Reproducible Research Peer Assessment 1
========================================================
```{r computetime,echo=FALSE}
time <- format(as.POSIXlt(Sys.time(), "GMT"), "%a %b %d %X %Y")
```
This markdown file was generated at `r time` UTC.

## Introduction

## Set Parameters and Load Data


First the working directory was set

```{r globalparameters, results="hide"}
setwd("~/GitHub/RepData_PeerAssessment1-1")
Sys.setlocale("LC_TIME", "English")
require(data.table)
require(ggplot2)
options(scipen = 1, digits = 7)

```


Then, we will load in and preprocess the data.

```{r loaddata}
fileUrl <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
if(!file.exists("activity.csv")){
        download.file(fileUrl, destfile="activity_monitoring_data.zip", 
                      method="auto")       
        unzip("activity_monitoring_data.zip")
}
activityData <- data.table(read.csv("activity.csv",header = TRUE, sep = ",",
        na.strings = "NA", colClasses=c("numeric","Date","numeric")))
```

## What is mean total number of steps taken per day?

Next we will determine what is the mean number of steps taken per day.

```{r}
transformedDataByDate <- activityData[,list(total=sum(steps)),by=date]
stepsMean <- mean(transformedDataByDate$total,na.rm=TRUE)
stepsMedian <- median(transformedDataByDate$total,na.rm=TRUE)
ggplot(transformedDataByDate, 
        aes(x=total)) + geom_histogram(color="black", fill="white", binwidth = 1000) +  
        geom_vline(aes(xintercept=stepsMedian), color="red", linetype="dashed",
        size=1) + ylim(0, 18) + ggtitle("Total Number of Steps Taken Per Day: Missing Data Ignored") + xlab("Steps per day") + ylab("Count") + theme_bw()
```

The mean number of steps taken per day is `r stepsMean`.
The median number of steps taken per day is `r stepsMedian`.

## What is the average daily activity pattern?

Next sought to determine the average daily activity pattern.

```{r}
transformedDataByInterval <- activityData[,list(average=mean(steps,na.rm=TRUE)),
        by=interval]
ggplot(transformedDataByInterval,aes(x=interval,y=average))+ geom_line() + ggtitle("Average Daily Activity Pattern")+ theme_bw()
maxInterval<-transformedDataByInterval[which.max(transformedDataByInterval$average),
        interval]
```

The 5-minute interval, on average across all the days in the dataset, containing 
the maximum number of steps is `r maxInterval`.

## Determing the effect of imputing missing values

```{r}
numberNA <- length(is.na(activityData$steps)[is.na(activityData$steps)==TRUE])
imputedActivityData <- activityData
imputedActivityData$steps <- ifelse(is.na(activityData$steps), 
        transformedDataByInterval[match(activityData$interval, 
        transformedDataByInterval$interval),transformedDataByInterval$average], 
        activityData$steps)
```

In the original data set, there exists `r numberNA` missing values in the dataset.
As a result, missing values were imputed by substituting the the mean for that 5-minute interval across all days for the missing value.

```{r}
transformedImputedDataByDate <- imputedActivityData[,list(total=sum(steps)),
        by=date]
stepsImputedMean <- mean(na.omit(transformedImputedDataByDate$total))
stepsImputedMedian <- median(na.omit(transformedImputedDataByDate$total))
ggplot(transformedImputedDataByDate, aes(x=total)) + geom_histogram(color=
        "black", fill="white",binwidth = 1000) + geom_vline(aes(xintercept=stepsMedian), 
        color="red", linetype="dashed", size=1) + ylim(0, 18) + ggtitle("Total Number of Steps Taken Per Day: Missing Data Imputed") + xlab("Steps per day")+ ylab("Count")  + theme_bw()
dataHis1 <- transformedDataByDate
dataHis1$datasource <- "Remove Missing"
dataHis2 <- transformedImputedDataByDate
dataHis2$datasource <- "Impute Missing"
dataHisBoth <- rbind(dataHis1,dataHis2)
ggplot(dataHisBoth, aes(x=total, fill=datasource)) + geom_histogram(color="black", binwidth=1000, alpha=.5, position="identity")+ xlab("Steps per day")+ ylab("Count")   + theme_bw() + theme(legend.position = c(0.875, 0.9),legend.background=element_rect(color = "black", size = 1, linetype = "solid"))
```

Changes in mean and median as result of imputation were negligible.  For example,
the preimutation mean and median were `r stepsMean` and `r stepsMedian`, 
respectively, while the postimputation mean and median were `r stepsImputedMean`
and `r stepsImputedMedian` respectively.

If we plot both preimputed and postimputed data together, we can see that the bar in the histogram corresponding to the postion

## Are there differences in activity patterns between weekdays and weekends?

Next, we sought to determine whether differences in the activity pattern existed
between weekdays and weekends.

```{r}
dataframe <- imputedActivityData
dataframe$weekdayorweekend <- ifelse(weekdays(dataframe$date) %in% c('Saturday',
        'Sunday'),"weekend", "weekday")
SHIT <- dataframe[,list(average=mean(steps)),by=list(interval,
        weekdayorweekend)]
ggplot(SHIT,aes(x=interval,y=average))+ geom_line()+ facet_grid(weekdayorweekend~.) +ggtitle("Average Daily Activity Pattern: Weekday vs. Weekend") + theme_bw()
```



